The Cost of Exclusion: Political Discrimination and Trust Deficits in the European Union

### 2025-07-09 article/main.tex
New text

Supporting this view, Levy and Sacks (2009) show that perceived government unresponsiveness among disadvantaged groups is associated with lower trust and political efficacy. Anderson and Tverdova (2003) and Oskooii (2024) similarly find that perceived injustice, whether interpersonal or institutional, generates cognitive and emotional responses that undermine confidence in state institutions.

While this framework has typically focused on ethnic or cultural exclusion (Wilkes, 2018; Tyrberg, 2023), political exclusion—based on one's beliefs or partisan identity—has received comparatively less attention. Yet, discrimination or harassment for political reasons may function in similar ways. Such experiences act as signals of exclusion from public discourse or decision-making, leading individuals to feel marginalized from the political community. This sense of exclusion can cause individuals to question the impartiality, responsiveness, and integrity of institutions.

@@

Political dissonance is exacerbated in polarized societies where outgroup hostility, moral condemnation, and dehumanization are common (Finkel et al., 2020). This intensifies the emotional toll of political discrimination and heightens perceptions of institutional betrayal (Hetherington \& Rudolph, 2015). In turn, this reduces trust and discourages political participation (Michelson, 2003; Lelkes \& Westwood, 2017).

@@

The second mechanism concerns the politicization of institutional perception—the tendency to evaluate institutions based not on their procedures or performance, but on perceived political alignment. In highly polarized contexts, even neutral institutions are judged through partisan filters (Rogowski \& Sutherland, 2015), leading to a fragmentation of legitimacy. Trust becomes conditional on whether individuals perceive institutions as aligned with their own political views.

@@

This asymmetric trust dynamic reflects the erosion of institutions as shared reference points. Studies show that even traditionally impartial bodies—courts, electoral commissions, public broadcasters—become objects of distrust when seen as favoring one side (Druckman \& Levendusky, 2019; Keefer \& Scartascini, 2022). As Oskooii (2016) notes, perceived exclusion reshapes institutional evaluations, reframing institutions as structurally biased rather than neutral arbiters.

>>>>

Supporting this view, Levy and Sacks (2009) show that perceived government unresponsiveness among disadvantaged groups is associated with lower trust and political efficacy. Anderson and Tverdova (2003) and Oskooii (2016) similarly find that perceived injustice, whether interpersonal or institutional, generates cognitive and emotional responses that undermine confidence in state institutions.

While this framework has typically focused on ethnic or cultural exclusion (Oskoii, 2016; Wilkes & Wu, 2019; Tyrberg, 2023), political exclusion—based on one's beliefs or partisan identity—has received comparatively less attention. Yet, discrimination or harassment for political reasons may function in similar ways. Such experiences act as signals of exclusion from public discourse or decision-making, leading individuals to feel marginalized from the political community. This sense of exclusion can cause individuals to question the impartiality, responsiveness, and integrity of institutions.

@@

Political dissonance is exacerbated in polarized societies where outgroup hostility, moral condemnation, and dehumanization are common (Iyengar & Westwood, 2015). This intensifies the emotional toll of political discrimination and heightens perceptions of institutional betrayal (Hetherington \& Rudolph, 2015). In turn, this reduces trust and discourages political participation (Michelson, 2003; Lelkes \& Westwood, 2017).

@@

The second mechanism concerns the politicization of institutional perception—the tendency to evaluate institutions based not on their procedures or performance, but on perceived political alignment. In highly polarized contexts, even neutral institutions are judged through partisan filters (Rogowski \& Sutherland, 2015; Druckman \& Levendusky, 2019), leading to a fragmentation of legitimacy. Trust becomes conditional on whether individuals perceive institutions as aligned with their own political views.

@@

This asymmetric trust dynamic reflects the erosion of institutions as shared reference points. Studies show that even traditionally impartial bodies—courts, electoral commissions, public broadcasters—become objects of distrust when seen as favoring one side (Keefer \& Scartascini, 2022). Lenz (2012) demonstrates that voters often adjust their perceptions of institutional performance based on partisan cues, even when confronted with neutral or contrary evidence. Similarly, Nyhan and Reifler (2010) find that political misperceptions persist because citizens interpret information through motivated reasoning, reinforcing distrust in institutions perceived as aligned with opposing political forces. These dynamics are particularly salient among individuals who already feel politically excluded: as Oskooii (2016) shows, such experiences of marginalization prompt individuals to reinterpret institutions as structurally biased, rather than as neutral arbiters of the democratic process.
